\textbf{\supplement}\vskip 2ex \mytitle\vskip 2ex

### 2022-02-28 paper-seb-main.tex
All files updated for PLOS Feb 28 2022 submission

\renewcommand\thesubsection{(\thesection.\alph{subsection})}
\usepackage{hyperref}

\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}

\makeatletter
    \def\do@url@hyp{\do/}
    \usepackage{moreverb,url}
    %    \renewcommand{\thefootnote}{\arabic{footnote}}
\makeatother

@@

\begin{document}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Harold Thimbleby
\\
\bigskip
See Change Fellow in Digital Health
\\
\bigskip

% Current address notes

@@

\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.

\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.

@@

%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.
\linenumbers

%\maketitle

@@

We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained. An interesting overlap with statistical and software engineering scrutiny is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA)\@.'' yet \emph{how\/} those analyses might have been performed is not discussed. The problem is that admitting using SAS 9.2 or any other named system does not really help scrutiny, as such systems can do anything with the data. A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is reliable.

@@

Contrary to the views expressed in the present paper, some authors have asserted that code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality}. On the contrary, if code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways. Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.

@@

@@

It should be noted that almost all code contains ``magic numbers'' --- that is, data masquerading as code. This common practice ensures that published data is very rarely all of the data because it omits the magic numbers. This emphasizes the need for repositories to require the inclusion of code so all data is actually available.

Bugs can be understood as discrepancies between what code is intended to do and what it actually does. Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered. Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination. None of the code examined for this paper includes either.

If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs. Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things. For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the sense of producing incorrect results --- that is, what was wanted was wrong. Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.

\iffalse Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, unqualified programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering. Dunning and Kruger go on to say,

@@

It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}. We \emph{all\/} make mistakes we are unaware of. If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified.

@@

Computers themselves are also a part of the problem. Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable. Programs can be written so that it is not possible to determine what they do or how they do it (whether deliberate obfuscation or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}. The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
\fi

@@

Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021). It is
\newcount\u
\newcount\v
\newcount\w
\newcount\frac
\input generated-covidsimSummary.tex
kLOC written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done. If nothing else, as the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models. It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place. The revised code has now been published, and it has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.

@@

\subsection{Science beyond epidemiology}
Code needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben} that, if not articulated, can affect results in unknown ways that may undermine any claims. People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place. The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even downright flaws.

@@

\begin{table*}
\begin{center}\normalsize
\input generated-summary-table.tex
%\IfFileExists{generated-summary-table.tex}{\input{generated-summary-table.tex}}{Missing file\typeout{No generated-summary-table.tex - error ignored, but you need to latex paper-seb-supplementary-material.tex to generate it}}
\end{center}

@@

\input generated-clone-date.tex
\begin{center}

@@

\input generated-repos.tex
\vskip 1ex
\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
\\
\small Repository clones downloaded and automatically summarized \clonedate
\end{center}
\newcount \pubdelayinmonths
\pubdelayinmonths=\cloneyear
\advance \pubdelayinmonths by -2020
\multiply \pubdelayinmonths by 12
\advance \pubdelayinmonths by -4 % published in April
\advance \pubdelayinmonths by \clonemonth
\caption{Sizes of repositories, with approximate sizes of code (in kLOC, thousands of lines of code) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison. Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting. Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper. Paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ has nothing in its repository except a single file still saying ``Code coming soon...'' despite, at the time of cloning and checking repositories, \the\pubdelayinmonths\ months had already elapsed since the paper referring to the code was published.}
\label{table-repo-summary}
\end{table*}

\section{A pilot survey of peer-reviewed research relying on code}
\label{survey-section}
The problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. For instance, a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model} provides no details of its code. Its \supplement, which might have provided code, no longer exists.

Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized. We therefore undertook a survey of recent papers in leading peer reviewed journals.

\subsection{Methodology}
A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected published in the leading journals \journalBreakdown.

The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal. All papers sampled are Open Access, although for some papers some or all of the associated data has restricted access. Table \ref{table-overview} is an overview of the sample.

Papers were selected from the journals' July 2020 new online listings where the paper's title implied that code had been used in the research. Commentary, correspondence, and editorials were excluded. The sample is intended to be large enough and objective enough to avoid the selection bias in the papers that motivated the current paper (the sample excludes the motivating papers discussed above as they were not published in the sampled journals), so that the sample may be considered to fairly represent what the editorial and the broader peer review community in leading journals considers to be good practice for computationally-based science.

The selection criterion selected papers where the title implies the authors themselves considered code to be a significant component of the scientific contribution, and, indeed, all sampled papers relied on code in their research.

The sample may be considered to be small given the importance of the research questions and relative to the diversity and huge number of scientific papers, but (\emph{a\/}) the selected journals are clearly leading scientific journals that conform to and set the standards for scientific publishing practice more generally, and (\emph{b\/}) as will be clear from the following discussion, there is little variation across the sample, which implies that a larger sample would not have been usefully more insightful.

%The survey was not intended to be a formal, representative sample of all scientific research in general, but it is hoped it will be sufficient to dispel the possibility that the issues described above in earlier sections of this paper are isolated practice, perhaps an idiosyncrasy of a few authors or of a particular field, or perhaps due to a chance selection bias (e.g., the Ferguson papers were reviewed above because of Ferguson's public profile and the importance of dependable pandemic research, but they somehow might have just happened to be software engineering outliers).

>>>>

@@

\begin{document}
\linenumbers

@@

% Title must be 250 characters or less.
\begin{flushleft}
\begin{change}
{\Large
\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\end{change}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Harold Thimbleby
\\
\bigskip
See Change Fellow in Digital Health
\\
\bigskip

@@

\begin{change}\noindent\emph{Methodology:} This paper critiques the quality of software engineering practice, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions, and undertakes a pilot survey of peer reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.
\end{change}

\noindent\emph{Results:} Scientists rarely assure the quality of code they rely on, rarely make complete code available, and rarely provide adequate documentation to understand or use their code reliably. \begin{change}Journals do not always have a code policy, and many have a relaxed approach.\end{change}

@@

%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.

@@

We would not believe a statistical claim that was obtained from some \emph{ad hoc\/} analysis with a new fangled method devised just for one project --- instead, we demand statistics that is recognizable, even traditional, so we are assured we understand what has been done and how reliable results were obtained. An interesting overlap with statistical and software engineering sloppiness is the many papers that just disclose as part of their methodology that they used a particular statistical package (e.g., ``Data analyses were performed using SAS 9.2 (SAS Institute, Cary, North Carolina, USA),'' yet \emph{how\/} those analyses might have been performed is not discussed. The problem is that admitting using SAS 9.2 or any other named system does not help scrutiny, as such systems can do anything with the data. A reviewer, if nobody else, needs to actually examine the statistical code itself and its documentation to assess whether the analysis presented in the paper is appropriate and suffciently reliable.

@@

Contrary to the views expressed in the present paper, some authors have asserted that the purpose of code is to provide insight into models, rather than precise (generally numerical) analyses summarising data \cite{assessing-quality} --- code can also be used to analyze and critique scientific theories directly. If code is inadequate, ``insights'' it provides will be flawed, and flawed in unquantified and unknown ways. Indeed, none of the papers sampled (described below in section~\ref{survey-section}) claimed their papers were using code for insight; all claimed, explicitly or implicitly, that their code outputs were integral to their peer reviewed results.

@@

\begin{change}
\subsection{The normal scientific emphasis on data}
Data has been at the center of science, certainly since the earliest days of astronomy collecting planetary and other information. Today it is widely recognized that lack of accessible and usable data that has already been collected limits the progress of science. Low quality data and poor access to data causes reproducibility problems, an increasingly recognized problem --- in 2015 it was estimated that \$28B a year is spent on preclinical research alone that is not reproducible \cite{preclinical-reproducibility}.

There are many current activities to proceduralize and standardize the more effective curation and use of data, such as the FAIR principles (Findable, Accessible, Interoperable and Reusable --- both for machines and for people) for scientific data management and stewardship \cite{fair}, and in the development of journal and national funder policies. For example, the 2022 update to the US National Institutes of Health data (not code) policies \cite{nih-policy} is described as a ``seismic mandate'' by \emph{Nature\/} \cite{nih-nature} in its attempt to improve reproducibility and open science.

On the whole, these cost estimates and initiatives under-play or ignore the role of software as a critical part of at least some scientific endeavor: code has become the new laboratory for almost all science. The role of code specifically in modeling is discussed throughout this paper; without bespoke code, proposed models (unless very abstract) cannot make a quantifiable contribution to the literature. More generally, much data is embedded in code, and in the limit code and data are indistinguishable (see \supplement). Furthermore, code has additional problems of versions and compatibility beyond those of data, for example suitable compilers to run old code may no longer be available, and --- worse --- programming systems may silently produce different results when used on different computers.

In general, without proper management of code --- for example to detect and report version control differences --- sharing code may even be counter-productive. (The data and code shared with the present paper includes cryptographic checksums; if somebody reproducing the work described here does not obtain the same checksums, then there are problems that need investigation before relying on the reproducibility of the data.)

\subsection{The deceptive simplicity of code}\label{deceptive-simplicity-of-code}
It is a fallacy that programming is easy \cite{fixit}. More correctly, toy programming is easy, but real programming is very difficult.

An analogy helps. Building houses is very easy --- indeed, many of us have built toy Lego houses. Obviously, though, a Lego house is not a \emph{real\/} house. It is not large enough or strong enough for safe human habitation! This is obvious because we can see Lego houses. In contrast to Lego, computer programs are generally invisible, and therefore the engineering problems within them are also invisible. Thus the ``programming is easy'' clich\' e is deceptive --- programming appears easy \emph{because\/} professional standards of building software are ignored, because people cannot see the reasons why they are needed. It is like sincerely appreciating a child's Lego building because we are not worried about subsidence, load bearing, electric shock, fire risks, water ingress, or even planning regulations.

Certainly, building is much easier and faster when such technical details are ignored, as anyone who has experienced a cowboy build may attest.

Unlike building (the Code of Hammurabi dates to around 1755{\sc bc}), programming is a very new discipline, and the problems of poor programming are not widely appreciated. Relevant professional standards are not enforced. Problems for the reliability of science arise when legitimate doodling with software drifts into claiming scientific results that do not have the reliable engineering structures underpinning them, let alone the properly developed and documented archived code, to justify them. In many countries, you cannot even start to build without first having plans approved, but who writes plans for software?

Since programming appears to be so easy, developing code has low status. The developers of code are rarely acknowledged in scientific papers. The implicit reasoning is: if programming is easy, then its intellectual contribution to science is negligible, so it is not even worth citing it or acknowledging the technical contributors to it. While this view prevails, the vicious cycle is that the low status means software development is done casually, which reinforces the low status.

In reaction to this vicious cycle, there is a growing movement to cite code correctly \cite{cite-code}, because code \emph{is\/} important, particularly for reproduction, testing and extension of any scientific work.

Scientists have been collecting and analyzing data for millennia, and curating data is taken seriously as a part of normal science and peer reviewed publication. Journal policies widely require appropriate discussion of data (much like they require appropriate discussion of statistics). Journals often require archiving data in standard formats so it can be accessed for further scientific work, such as reproduction.

Unfortunately, very few scientists and even fewer journals realize that data and code are theoretically and in practice indistinguishable (see \supplement). Given that data and code are equivalent and interchangeable, it follows that publishing policies on data handling --- and the reasons for those policies --- should also apply at least as strictly to code.

\subsection{The central role of code is ignored}\label{central-role-of-code}
It is important that experiments and analysis, such as statistical analysis, are performed reliably and ethically. There are many protocols and journal policies that enforce good practice, for example a journals often require adherence to PRISMA (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) \cite{prisma} for any paper performing a systematic review of the literature. Yet PRISMA, like many policies, ignores the central role of computers, and ignores the software engineering principles that assure computation is reliable and reliably reported.

There are many more styles of scientific investigation than systematic reviews, of course, but as PRISMA has had considerable development and is widely required by journals, it will serve as an example of overlooking the benefits of code in doing science.

PRISMA ``was designed to help systematic reviewers transparently report why the review was done, what the authors did, and what they found,'' which sounds reasonable enough. PRISMA covers the review process carefully. For example, the authors should report the number of papers they included in their review. Perhaps $N=1093$. This number is then written into their draft paper, likely in several places. As the authors reads and revises their paper, submit it and respond to peer reviewers, it is likely that the number of papers included in the survey changes, or other details may change too. The authors now have a maintenance problem: where are the numbers that have changed, and what should they now be changed to? Doing a search-and-replace, whether automated or by hand, is fraught with difficulties. What happens if $1093$ is used for some other purposes as well? What happens if some of the $1093$ values were written as $1,093$ and are not noticed? Then there are the Human Factors, where slips and errors will happen in this process anyway. Similar iterative revision cycles happen with any paper, not just with systematic reviews. Typos, slips during cut-and-paste, and other errors are common. They are problems across all science.

If a computer program is involved in the process (it generally is for a systematic review) then the value of $N$ could easily be stored in a file where the paper typesetting process can access it. For example, if \LaTeX\ is the system of choice, the analysis could generate, say,

\begin{center}\texttt{$\backslash$newcommand\{$\backslash$N-papers-reviewed\}\{1093\}}\end{center}

so that when and wherever the author writes \texttt{$\backslash$N-papers-reviewed} in their paper's text, the typeset paper says 1093 or whatever the correct value happens to be at the time. If so, then \emph{whenever\/} the survey is updated, the value cited in the paper is \emph{correctly\/} updated without any further intervention from the author.

In general, not just numbers, but any data, text, graphs or tables, etc, from the analysis can be reliably inserted into a paper automatically.

A special case of this process is the automatic handling of bibliographies and citations: what used to be a tedious manual process is now very commonly automated and for exactly the same reasons, that is, simultaneously improving ease of use and reliability. Note that automatic bibliographies insert both free text and citation numbers: data values are arbitrary. Moreover, automatic bibliography systems also provide many useful validation checks on the data, for instance that a journal article citation actually specifies a journal name.

Increasingly, many analysis systems have themselves introduced features to automate keeping papers up to date and for reliably referring to correct values (what Software Engineers would call maintaining invariants with the data). R~(e.g., using RMarkdown) and Mathematica are popular examples of such general purpose analysis systems that provide powerful features to help authors more easily produce reliable papers. In particular, they make it completely trivial to read and edit a paper, revise the analysis, redraw tables and diagrams etc, and obtain a correctly revised paper with no further manual work. Crucially, this encourages the author to explore the analysis and competing analyses more than usual, so their confidence in the final analysis and quoted results is much higher than usual. Furthermore, RMarkdown and Mathematica also have powerful mechanisms to create slide shows, and therefore the guarantees of automatic correctness transfer to presentations about the research as well.

An important consequence of using a code-based automated process is that it removes barriers to revising and improving papers. Before such methods were used, it was tedious and risked introducing new errors if one undertook to update research and papers. When the process is automated, however, there are no such barriers.

Of course errors may still occur, but there is an important and dramatic improvement. Each time a paper is processed, the \emph{same\/} code is re-used. So each revision and proof reading of the paper helps detect those errors, and once any error is detected and corrected it will not arise again. Moreover, since the same potential error, if any, is repeatedly proof read and checked (it occurs in each iteration of the paper, and perhaps in many places in each version of the paper), this reduces the chances that the error is missed. This is a very different situation to the manual processing of data, where every time a paper is edited or revised, the data may be corrupted accidentally in new ways and in new places.

The present paper performed a \ifnum \dataN<100 \textbf{small}\fi\ pilot survey of \textbf{\plural{\dataN}{paper}} sampled from \textbf{\plural{\numberOfJournals}{journal}}, \textbf{\the\counthasDataRepository} of which used code repositories. The repositories were most recently accessed on \textbf{\clonedate} for automatic analysis for generating the summary data reported in this paper. All those facts, highlighted in bold here but also used throughout this paper and in the \supplement, are ensured correct by the author using an automatic process (discussed in full in the \supplement). One should note, as emphasized above, that as these specific values as used above also appear in many places and in many different contexts in the present paper, this significantly reduces the Human Factors problems of noticing errors.\footnote{\begin{change}For example, the author types 1093, but makes a transposition slip and writes 1039. Since the author thinks they know what they wrote, when they proof read 1039 they misread it (this is a very common problem with misspellings and grammatical ambiguities). These errors (transposition and misreading, etc) do not happen consistently each time, so the more often 1039 occurs in the paper \emph{if generated automatically}, the more likely the errors will be addressed. See \cite{enigma} for an actual example.\end{change}}

PRISMA says nothing about how to ensure the final results of a survey are correctly and reliably presented in a paper, despite this being one of PRISMA's explicit motivations. Ironically, many journal publishing policies (including \emph{PLOS ONE}\/) forbid using such techniques (see the \supplement\ for further discussion). Again, such rules reinforce the fallacy that code is trivial and unimportant. 
\end{change}

@@

It should be noted that almost all code contains ``magic numbers'' --- that is, data masquerading as code. This common practice ensures that published data is very rarely all of the data because it omits the magic numbers. This emphasizes the need for repositories to require the inclusion of code so all data, \begin{change}including that embedded in the code\end{change}, is actually available.

Bugs can be understood as discrepancies between what code is ought to do and what it actually does. Many bugs cause erroneous results, but bugs may be ``fail safe'' by causing a program to crash so no incorrect result can be delivered. Better, contracts and assertions are essential defensive programming technique that block compilation or block execution with incorrect results; they turn bugs into safe termination. None of the code examined for this paper includes any such techniques.

If code is not documented it cannot be clear what it is intended to do, so it is not possible to detect and eliminate bugs. Indeed, even with good documentation, \emph{intentional bugs\/} will remain, that is, code that correctly implements the wrong things \cite{essence-of-software,fixit} --- they are bugs that were intended but were ideas based on mistaken ideas (students and inexperienced programmers make intentional bugs all the time). For instance, in numerical modeling, using an inappropriate method can introduce errors that are not ``bugs'' in the narrow sense of incorrectly implementing what was wanted (e.g., ill-conditioning), but are bugs in the wider sense of producing incorrect results --- that is, what was intended was wrong.

Random numbers are widely used in computational science, e.g., for simulation or for randomizing experiments. Misuse of random numbers (e.g., using standard libraries without testing them) is a common cause of bugs \cite{knuth}.

\begin{change}
\subsection{Human Factors of programming}
People would generally not make coding errors if there were aware they were making errors. Unfortunately programming is a very demanding activity, which causes tunnel vision (also known as loss of situational awareness). The consequence is programmers focus on ``the'' problem as it appears in the code, and are unaware they are not considering wider issues. The correctness, generality, and usability of a program are therefore often unintentionally sacrificed to making code work at all.

Confirmation bias is a standard Human Factors problem \cite{fixit} (which scientists have many techniques to avoid), which encourages us to perform tests that show our programs work. Instead, we should be rigorously testing ways in which programs can fail. This is exactly the same issue pointed out by Popper \cite{popper}: scientists should experiment to find reasons why hypotheses are false, and indeed use simple hypotheses that are testable. Software is really no more than a collection of sophisticated hypotheses.

Standard Human Factors mitigations for these problems include team working, with appropriate precautions to manage authority gradients (where the Human Factors oversights of the leader influence the team). Many computerized mitigations are also available --- strong typing, code analyzers, formal methods, and so on. See the \supplement.

Following the Dunning-Kruger Effect \cite{dunning-kruger,dunning-kruger-numeracy}, programmers over-estimate their programming skills because they do not have the skills to recognize their lack of knowledge, specifically in the present case, knowledge of basic software engineering. Dunning and Kruger go on to say,

@@

It sounds harsh to call programmers incompetent, but challenged with the complexity of programs and the complexity of the domains programs are applied in, we are all incompetent and succumb to the limitations of our cognitive resources, suffering blindspots in our thinking \cite{fixit}. We \emph{all\/} make mistakes we are unaware of. If we do not have the benefit of professional qualifications that have assessed us objectively, we generally have a higher opinion of our own competence than is justified. Moreover, if we do not work in a diverse team, nobody will ever point this out, so the potential problems it causes will never be addressed.

@@

Computers themselves are also a part of the problem. Na\"\i vely modifying a program (as may occur during debugging) typically makes it more complex, more \emph{ad hoc}, and less scrutable. Programs can be written so that it is not possible to determine what they do or how they do it (whether by deliberate obfuscation, as in malware, or accidentally), except by running them, if indeed it is possible to exactly reproduce the necessary context to do so \cite{framework}. The point is, introducing bugs should be avoided so far as possible in the first place, and programs should routinely have assertions and other methods to detect those bugs that are introduced (see this paper's \supplement\ for more discussion of standard programming methodologies).
\end{change}

@@

Because of the recognized importance of the Ferguson paper, a project started to document its code  \cite{refactoring}.\footnote{The system is open source, available at \url{github.com/mrc-ide/covid-sim} version (19 July 2021). It is
\newcount\u
\newcount\v
\newcount\w
\newcount\frac
\input generated-covidsimSummary.tex
kLOC (thousands of lines of code) written in C++ with Python, R, sh, YML/JSON, etc, not C\@.}
Documenting the code now may describe what it does, \emph{including\/} its bugs, but it is unlikely to explain what it was intended to have done. If nothing else, as the code is documented, bugs will be found, which will then be fixed (refactoring), and so the belatedly-documented code will not be the code that was used in the published models. It is well-known that documenting code helps improve it, so it is surprising to find an undocumented model being used in the first place. The revised code has now been published, and it has been heavily criticized \citeeg{bad-code}, supporting the concerns expressed in the present paper.

@@

\section{Science beyond epidemiology}
\label{section-science-beyond-epidemiology}
\begin{change}
Epidemiology has a high profile because of the current COVID pandemic. There are many other areas of computational science that are equally if not more critical, and many will have longer-lasting impact. Climate change modeling is one such example.

A short 2022 summary of typical problems of software engineering impacting science generally appears in \emph{Nature} \cite{nature-review}, describing diverse and sometimes persistent problems encountered during research in cognitive neuroscience, psychology, chemistry, nuclear magnetic resonance, mechanical and aerospace engineering, genomics, oceanography, and in migration. The paper \cite{nature-review} unfortunately makes some misleading comments about the simplicity of software engineering, e.g., ``If code cannot be bug-free, it can at least be developed so that any bugs are relatively easy to find.''

Guest and Martin \cite{psychological-modeling} in another 2022 paper promote the use of computational modeling, arguing that through writing code, one debugs scientific thinking. Psychology, Guest and Martin's focus, has an interesting relationship with software, as computational models are often used to model cognition and to compare results with human (or animal) experiments \cite{psychological-modeling}. In this field, the computation does not just generate results, but is used to explicitly explore the assumptions and structures of the scientific frameworks from which the models are derived. Computational models can even be used to perform experiments that would be unethical on live participants, for instance involving lesioning (damaging) artificial neural networks. It should be noted that such use of cognitive models is controversial --- on the one hand, the software allows experiments to be (apparently) precisely specified and reproduced, but on the other hand in their quest for psychological realism the models themselves have become very complex and it is no longer clear what the science is precisely (e.g., ACT-R, one widely-used theory for simulating and understanding human cognition, has been under development since 1973 and is now a 120 kLOC Common LISP and Python system \cite{actr}; and of course any paper using ACT-R would require additional code on top of the basic framework).

The psychology paper \cite{psychological-modeling} demonstrates building an example computational model from scratch to illustrate their own framework of computational science. In fact their example model has no psychological content: a simple numerical test is performed, but the psychology of why the result is counterintuitive --- the psychological content --- is not explored. Be that as it may, they develop a mathematical specification and discuss a short Python program they claim implements it.

The Python code is presented without derivation or discussion, as if software engineering is trivial. The program listed in the paper certainly runs without obvious problems (ignoring typographical errors due to the journal's publishers), but ironically the Python does not implement the mathematical specification explicitly provided for it,\footnote{\begin{change}More precisely: the program has a bug, and/or the specification given is wrong or too abstract.\end{change}} thus unintentionally undermining the argument of the paper.

One might argue the bug is trivial (the program prints \texttt{False} when it should print \texttt{b}), but to dismiss such a bug would be comparable to dismissing a statistical error that says $p=\mbox{\tt False}$ which would be nonsense --- if a program printed that, one would be justified in suspecting the quality of the entire program and its analyses. Inadvertently, it would seem, then, that the paper shows that just writing code does not help debug scientific thinking: instead, code must first be derived in a rigorous way and actually be correct (at least when finished). Otherwise, computational modeling with inadequate software engineering will very likely introduce errors into scientific thinking.
\end{change}

Code generally for any field of scientific modeling needs to be carefully documented and explained because all code has tacit assumptions, bugs and cybersecurity vulnerabilities \cite{Ben,nature-review,se-bias} that, if not articulated \emph{and properly managed}, can affect results in unknown ways that may undermine any claims. People reading the code will not know how to obtain good, let alone better results, because they do not know exactly what was intended in the first place. The problem is analogous to the problem of failing to elaborate statistical claims properly: failure to do so suggests that the claims may have unknown limitations or even downright flaws.

@@

\begin{table*}
\begin{center}\normalsize
\input generated-summary-table.tex
\end{center}

@@

\begin{center}

@@

\input generated-repos.tex
\vskip 1ex
\small Citation numbers $>$ \MaxMainPaperCitationNumber\ can be found in the \supplement
\\
\small Repository clones downloaded and automatically summarized \clonedate
\end{center}

\caption{Sizes of repositories, with approximate sizes of code (in kLOC) and data for all available GitHub repositories reviewed in the survey, plus \texttt{covid-sim} \cite{ICmodel} for comparison. Sizes are approximate because in all repositories code and data are conceptually interchangeable (an issue explained in the \supplement), so choices were made in the survey to avoid double-counting. Many repositories rely on downloading additional code and data, which is not counted in the table as the additional required material is not in the repository cited in the paper.  At the time of cloning and checking all repositories in \clonewhen, paper \csname cite-PostoperativeOutcomes.RiskNet\endcsname\ still had nothing in its repository except a single file still saying ``...code coming soon...,'' despite \the\pubdelayinmonths\ months having already elapsed since the submitted paper had claimed the code could be accessed in its repository.}
\label{table-repo-summary}
\end{table*}

\subsection{A pilot survey of peer-reviewed research relying on code}
\label{survey-section}
The problems of unreliable code are not limited to COVID-19 modeling papers, which, understandably, were perhaps rushed into publication. For instance, a 2009 paper reporting a model of H5N1 pandemic mitigation strategies \cite{flu-model} provides no details of its code. Its \supplement, which might have provided code, no longer exists.

Almost all scientific papers rely on generic computer code (for statistics or for plotting graphs, and so on) but what is of interest is whether, and, if so, to what extent, code developed \emph{specifically\/} as part of or to support a research contribution can be understood by colleagues and scientifically scrutinized. \begin{change}We therefore undertook a pilot survey of papers in leading peer reviewed journals.\end{change}

\subsubsection{Methodology}
\begin{change}
There are many ways to do good science. Normal science routinely involves either performing new research and reporting it in the literature, or performing reviews --- such as systematic reviews --- of the existing literature to form a consensus view of an existing body of research. Typically, research offers new theories or results, points out previous errors, or uses statistical or qualitative methods to reduce the variability in previously reported results. In any case, the intention is to provide reliable knowledge and to bolster consensus \cite{ziman}.

In the present case, we have identified limitations in the methodology of earlier scientific contributions, throwing serious doubt on their soundness and reliability; furthermore, we identified critical misunderstandings in the literature citing those papers, including misunderstanding the nature of code reproducibility. (Shortly, we will present some ways forward, in section \ref{section-discussion}.)

It is possible, if not likely (e.g., given the report in \emph{Nature\/}, cited above \cite{nature-review}), that such limitations are more widespread than the specific examples considered so far, which are anyway limited primarily to epidemiology, and to COVID research performed under extreme pandemic pressure. We now therefore undertake a small, manageable stratified sample of peer reviewed literature to explore whether the issues are more widespread.
\end{change}

A sample of \plural{\dataN}{recent paper} covering a broad range of science were selected. The papers were selected from the leading journals \journalBreakdown.

The two journals \emph{Nature Digital Medicine\/} and \emph{Lancet Digital Health\/} were selected as leading specialist science journals in an area where correctness of scientific modeling has safety-critical implications, and \emph{Royal Society Open Science\/} was selected as a leading general science journal. All papers sampled are Open Access, although for some papers some or all of the associated data has \begin{change}no or\end{change} restricted access, \begin{change}in some cases despite the relevant journal policies on code\end{change}. Table \ref{table-overview} is an overview of the sample.
